Scaling rates of true polar wander in convecting planets and moons

### 2017-05-01 tpw_rate.tex
Add a sentence on Cambiotti.

For Cenozoic Earth we can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch}.
Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
\citet{richards1997explanation} have argued that the convective planform of Earth has been stable for the last
few hundred million years. On the other hand, we know that there have been large reorganizations of that planform 
during Earth history, so this recent geologic stability may not hold in general \citep{evans2003true}.
Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
\begin{equation}
\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{\tau_R} \sim 6^\circ / \mathrm{Myr},
\end{equation}
which corresponds to about 66 cm/yr 90$^\circ$ from the TPW axis.
This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
which results in a corresponding uncertainty for the relaxation time $\tau_R$ and the maximum polar wander rate.

>>>>

We can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch} to obtain an estimate of maximum TPW rates for Earth.
Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
\citet{richards1997explanation} performed TPW simulations based on Cenozoic plate motion reconstructions, and argued that the convective planform of Earth has been stable for the last few hundred million years. 
\citet{cambiotti2011new} took a similar approach, but allowed for the position of the spin axis to lag the principal axes (i.e. nonzero $\theta$). In that study they found values for $\theta$ of up to 3-7$^\circ$.
On the other hand, we know that there have been large reorganizations of that planform during Earth history,
so this relative stability of the spin-axis may not hold in general \citep{evans2003true}.
Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
\begin{equation}
\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{\tau_R} \sim 6^\circ / \mathrm{Myr},
\end{equation}
which corresponds to about 66 cm/yr 90$^\circ$ from the TPW axis.
This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
which results in a corresponding uncertainty for the relaxation time $\tau_R$ and the maximum polar wander rate.
